---
title: "Reproducible Final Project"
author: "Carson Archie"
date: "2026-04-23"
output: html_document
---

Below are my graphs and data that I have generated for my final project.
This is the link to my GitHub Repository: https://github.com/CarsonArchie/Soil_MS_Carson.git 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we need to install and call on all of the packages we need along with the CSV's of load and ppm concentration data.
```{r packages and data}
#Install the necessary packages for running tests

#Load required packages 
library(pheatmap)
library(tidyverse)
library(dplyr)
library(ggcorrplot)
library(ggplot2)
library(lme4)
library(emmeans)
library(multcompView)
library(ggpmisc)
library(GGally)
library(lmodel2)
theme_set(theme_bw())

#Load in the required data from the working directory pulled from GitHub
load_raw_data <- read.csv("STP_LOAD_RAW(Sheet1).csv")
ppm_raw_data <- read.csv("STP_PPM_RAW(Sheet1).csv")
```


Below is a pairwise correlation plot with assigned values depending on the negative or positive corrlation between element species. 
```{r, correlation plot}
#Filter and order data for columns that will be analyzed
filter_data <- load_raw_data %>% 
  dplyr::select(Runoff_Vol:TSS_L)
#Keep only the numeric columns
filter_data <- filter_data %>% 
  dplyr::select(where(is.numeric))
#Compute correlation matrix to handle NAs properly
corr_matrix <- cor(filter_data, use = "pairwise.complete.obs")
if(any(is.na(corr_matrix))) {
  warning("Correlation matrix contains NAs. Consider removing columns with insufficient data.")
}
#Generate a plot using the correlation matrix
Correlation_Plot <- ggcorrplot(corr_matrix, 
                 hc.order = TRUE, 
                 type = "lower",
                 outline.col = "white",
                 ggtheme = theme_gray, 
                 colors = c("#6D9EC1", "white", "#E46726"), 
                 lab = TRUE,
                 title = "Runoff and Leachate Species Correlation")
print(Correlation_Plot)

```


This is a heat map which can show visual patterns of how runoff and leachate species interact and can have correlation. 
```{r, heatmap}

#Select numeric columns from 8 to 21 which contain all element species and TSS
numeric_cols <- names(load_raw_data)[8:21]
#Filter to keep only truly numeric columns (safety check)
numeric_cols <- numeric_cols[sapply(load_raw_data[numeric_cols], is.numeric)]
#Check if we have enough columns 
if (length(numeric_cols) < 2) {
  stop("Not enough numeric columns found in indices 8:21 to compute correlation.")
}
cat("Using columns for correlation:", paste(numeric_cols, collapse = ", "), "\n")
#Run a new pairwise correlation matrix
M <- cor(load_raw_data[, numeric_cols], use = "pairwise.complete.obs")
#Check for NA/NaN in correlation matrix
if (any(is.na(M))) {
  warning("Correlation matrix contains NAs. Some columns may have insufficient data.")
}
#Plot with Heat Map
#Save as a file(for some reason it wouldn't present if I didn't save it in a png)
pheatmap(M,
         color = colorRampPalette(c("#E69F00", "white", "#0072B2"))(100),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 9,
         fontsize_row = 8,
         fontsize_col = 8,
         main = "STP Load Data Correlations",
         filename = "STP_Correlation_Heatmap.png",
         width = 12,
         height = 10,
         dpi = 300)
cat("Heatmap saved to STP_Correlation_Heatmap.png\n")
#Display in RStudio Plot
pheatmap(M,
         color = colorRampPalette(c("#E69F00", "white", "#0072B2"))(100),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 10,
         main = "STP Load Data Correlations")
```


This gallery plot allows us to split correlations based on soil test phosphorus levels to get R^2 values and linear models along with concentration data to show the distribution and trends between STP and element species interations. 
```{r, gallery plots}
#Build out the upper function with equations to get your R^2 and regression equation
upperfun <- function(data, mapping){
  ggplot(data = data, mapping = mapping) +
    geom_blank() +
    stat_ma_eq(use_label(c("eq", "R2")),
               vstep = 0.15, 
               size = 2.7,
               formula = y ~ x)
}

#Build out the lower function ma lines using lmodel2 to get your correlations
lowerfun <- function(data, mapping){
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.4) +
    ggpmisc::stat_ma_line(show.legend =  F,
                          se = FALSE, 
                          method = "lmodel2:SMA")
}

#The diagonal function will show the distribution of species to 
diagfun <- function(data, mapping){
  ggplot(data = data, mapping = mapping) +
    geom_density(alpha = 0.33)
}
gallery_plot <- ggpairs(data = load_raw_data,
        columns = 8:21,
        aes(colour = Soil_P_Level_ppm, 
            fill = Soil_P_Level_ppm, 
            grp.label = abbreviate(Soil_P_Level_ppm)),
        legend = 1,
        upper = list(continuous = wrap(upperfun)),
        lower = list(continuous = wrap(lowerfun)),
        diag = list(continuous = wrap(diagfun)),
        labeller = as_labeller(function(x) {gsub("\\.", " ", x)}),
        progress = FALSE) +
  theme_bw(base_size = 9) +
  theme(panel.grid = element_blank())
gallery_plot
```


This is a summaries average version of the data from the load CSV which allows us to see trends in figues such as boxplots or line plots. 
```{r, create a mean summary load data table with columns for graphs}
summarized_load_mean <- load_raw_data %>%
  group_by(R_CSV_Label) %>%
  summarise(
#Keep specific non-numeric columns 
    Date = first(Date),  
    P_Rate_Kg.Ha = first(P_Rate_Kg.Ha),
    Replication = first(Replication),
    Soil_P_Level_ppm = first(Soil_P_Level_ppm),
    Label_ID = first(Label_ID),
    R_Group_ID = first(R_Group_ID),
    R_CSV_Label = first(R_CSV_Label),
#Calculate stats for numeric columns
    across(
      .cols = where(is.numeric), 
      .fns = list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        se = ~sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x)))
      ),
      .names = "{col}_{fn}"
    ),
    .groups = "drop"
  )

```


We have repeated this step for the ppm CSV. 
```{r, create a mean summary ppm data table with columns for graphs}
summarized_ppm <- ppm_raw_data %>%
  group_by(R_CSV_Label) %>%
  summarise(
#Keep specific non-numeric columns 
    Date = first(Date),  
    P_Rate_Kg.Ha = first(P_Rate_Kg.Ha),
    Replication = first(Replication),
    Soil_P_Level_ppm = first(Soil_P_Level_ppm),
    Label_ID = first(Label_ID),
    R_Group_ID = first(R_Group_ID),
    R_CSV_Label = first(R_CSV_Label),
#Calculate stats for numeric columns
    across(
      .cols = where(is.numeric), 
      .fns = list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        se = ~sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x)))
      ),
      .names = "{col}_{fn}"
    ),
    .groups = "drop"
  )


```



This is the coded function that creates stacked boxplots with color blind friendly display across all element species and faceted based on soil test phosphorus values. 
```{r stacked bar chart colorblind}
library(ggplot2)
library(dplyr)
library(tidyr)

#Define custom colors
cbbPalette <- c( "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#Reorder data levels for dates to make since
summarized_load_mean$Date <- factor(summarized_load_mean$Date, 
                                    levels = c("4/8/2025", "4/15/2025", "4/22/2025", "4/29/2025"))
generate_correlation_plots <- function(data, 
                                       x_var = "P_Rate_Kg.Ha", 
                                       y_vars_to_plot = NULL, 
                                       fill_var = "Date", 
                                       facet_var = "Soil_P_Level_ppm",
                                       plot_type = "bar", 
                                       fixed_yscale = TRUE,  
                                       output_prefix = "Plot") {
#Check Y variables if not provided
  if (is.null(y_vars_to_plot)) {
    exclude_cols <- c(x_var, fill_var, facet_var, "P_Rate_Kg.Ha_mean", "Replication_mean", "Label_ID_mean", "R_Group_ID_mean")
    all_nums <- names(data)[sapply(data, is.numeric)]
    y_vars_to_plot <- all_nums[grepl("_mean$", all_nums)]
    y_vars_to_plot <- setdiff(y_vars_to_plot, exclude_cols)
  }
  if (length(y_vars_to_plot) == 0) {
    stop("No numeric columns found to plot on the Y-axis.")
  }
  cat("Generating plots for", length(y_vars_to_plot), "variables.\n")
  plot_list <- list()
#Loop through each Y variable
  for (y_col in y_vars_to_plot) {
    nice_name <- gsub("_mean$", "", y_col)
    nice_name <- gsub("_", " ", nice_name)
    nice_name <- gsub("\\.", " ", nice_name)
  
    plot_data <- data
    if (plot_type == "bar") {
      grouping_vars <- c(x_var, fill_var, facet_var)
      grouping_vars <- intersect(grouping_vars, names(data))
      plot_data <- data %>%
        group_by(across(all_of(grouping_vars))) %>%
        summarise(y_val = mean(.data[[y_col]], na.rm = TRUE), .groups = "drop")
      geom_layer <- geom_col(position = "stack", width = 10, color = "black")
      
    } else {
      geom_layer <- geom_point(size = 3, alpha = 0.8, shape = 21, color = "black") +
                    geom_smooth(method = "lm", se = TRUE, linetype = "dashed", color = "gray40")
    }
#Calculate global Y max if using fixed scale
    y_limit <- NULL
    if (fixed_yscale && plot_type == "bar") {
      y_limit <- max(plot_data$y_val, na.rm = TRUE) * 1.1  
    }
#Determine Y column name based on plot type
    y_col_name <- ifelse(plot_type == "bar", "y_val", y_col)
#Build Plot
    p <- ggplot(data = plot_data, aes(x = .data[[x_var]], 
                                      y = .data[[y_col_name]], 
                                      fill = .data[[fill_var]])) +
      geom_layer +
#Add logic for scales
      facet_wrap(as.formula(paste("~", facet_var)), scales = ifelse(fixed_yscale, "fixed", "free_y")) + 
#Custom Colors
      scale_fill_manual(values = cbbPalette) + 
      labs(
        title = paste(ifelse(plot_type=="bar", "Stacked Bar:", "Correlation:"), nice_name, "based on P Rate"),
        x = "P Rate (Kg/Ha)",
        y = paste(ifelse(plot_type=="bar", "Mean", ""), nice_name, "Kg/Ha"),
        fill = "Date"
      ) +
      theme_minimal(base_size = 14) +
      theme(
        legend.position = "bottom",
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank()
      )
    
    plot_list[[y_col]] <- p
  }
  
  return(plot_list)
}
#Generate Stacked Bar Plots
my_bar_plots <- generate_correlation_plots(
  data = summarized_load_mean, 
  plot_type = "bar"
)
#View all plots
print(my_bar_plots)
# Save all plots
for (name in names(my_bar_plots)) {
  filename <- paste0(name, "_StackedBar.png")
  ggsave(filename, plot = my_bar_plots[[name]], width = 30, height = 6, dpi = 300)
  cat("Saved:", filename, "\n")
}
```


What is different regarding the ppm line plots is it allows for us to look as specific sample and test irregularities displayed in concentration beforw we move onto load data. This allows us to understand why there might be a low ppm, but a high delution due to lots of rainfall or vice versa. 
```{r line plots}
library(ggplot2)
library(dplyr)
library(tidyr)

# Define custom color palette
cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000") 

# Reorder data levels for dates
if(exists("summarized_ppm")) {
  summarized_ppm$Date <- factor(summarized_ppm$Date, 
                                levels = c("4/8/2025", "4/15/2025", "4/22/2025", "4/29/2025"))
}
generate_correlation_plots <- function(data, 
                                       y_vars_to_plot = NULL, 
                                       x_var = "Date",          
                                       group_var = "P_Rate_Kg.Ha", 
                                       facet_var = "Soil_P_Level_ppm",
                                       plot_type = "regression", 
                                       fixed_yscale = FALSE,     
                                       output_prefix = "Plot") {
  
#Auto check Y variables if not provided
  if (is.null(y_vars_to_plot)) {
    exclude_cols <- c(x_var, group_var, facet_var, "P_Rate_Kg.Ha_mean","Replication_mean", "Label_ID_mean", "R_Group_ID_mean")
    all_nums <- names(data)[sapply(data, is.numeric)]
    y_vars_to_plot <- all_nums[grepl("_mean$", all_nums)]
    y_vars_to_plot <- setdiff(y_vars_to_plot, exclude_cols)
  }
  
  if (length(y_vars_to_plot) == 0) {
    stop("No numeric columns found to plot on the Y-axis.")
  }
  cat("Generating plots for", length(y_vars_to_plot), "variables.\n")
  plot_list <- list()
#Loop over all applicable Y variables
  for (y_col in y_vars_to_plot) {
    nice_name <- gsub("_mean$", "", y_col)
    nice_name <- gsub("_", " ", nice_name)
    nice_name <- gsub("\\.", " ", nice_name)
#Convert the Group Variable to Factor
    plot_data <- data %>%
      mutate(
        .group_factor = as.factor(.data[[group_var]])
      )
    
#Calculate Y for a fixed scale
    y_limit <- NULL
    if (fixed_yscale) {
      y_limit <- max(plot_data[[y_col]], na.rm = TRUE) * 1.1}
    p <- ggplot(data = plot_data, 
                aes(x = .data[[x_var]], 
                    y = .data[[y_col]])) +
#Line Layer
      geom_line(aes(group = .group_factor, color = .group_factor), 
                linewidth = 0.8, alpha = 0.6) +
#Point Layer
      geom_point(aes(shape = .group_factor, color = .group_factor), 
                 size = 3, alpha = 0.9) + 
#Facet Layer
      facet_wrap(as.formula(paste("~", facet_var)), scales = ifelse(fixed_yscale, "fixed", "free_y")) +
#Scales
      scale_color_manual(values = cbbPalette, name = "P Rate (Kg/Ha)") +
      scale_shape_manual(values = c(16, 17, 15, 3, 4, 8, 9, 10), name = "P Rate (Kg/Ha)") +
      scale_fill_manual(values = cbbPalette, guide = "none") + 
#Labels
      labs(
        title = paste("Trend:", nice_name, "over Time"),
        subtitle = paste("Faceted by", facet_var),
        x = "Date",
        y = paste(nice_name, "(Mean)"),
        color = "P Rate\n(Kg/Ha)",
        shape = "P Rate\n(Kg/Ha)"
      ) +
      
#Theme
      theme_minimal(base_size = 14) +
      theme(
        legend.position = "bottom",
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )
#Apply Y limit if calculated
    if (!is.null(y_limit)) {
      p <- p + ylim(0, y_limit)
    }
    
    plot_list[[y_col]] <- p
  }
  
  return(plot_list)
}
#Generate Plots
my_reg_plots <- generate_correlation_plots(
  data = summarized_ppm, 
  plot_type = "regression",
  fixed_yscale = TRUE
)
#View all plots
if(length(my_reg_plots) > 0) {
  print(my_reg_plots)
}
#Save all plots
for (name in names(my_reg_plots)) {
  filename <- paste0(name, "_TimeTrend.png")
  ggsave(filename, plot = my_reg_plots[[name]], width = 12, height = 8, dpi = 300)
  cat("Saved:", filename, "\n")
}
```

