---
title: "Loops/FunctionChallenge"
author: "Carson Archie"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. The primary goal or point of writing your own functions and iterations  in R is too be able to have control over executing math equations and graphs without having human errors. It allows for precise and consistent replication across large data sets while ensuring that the same equation or process in being ran every time. 

2. 2.1 To write a function in R, it is good practice to store your designated function as an object using the <- opperation. Following this you will write the function() base opperator to build your desried function. You will classify what you are trying to do within the () brackets followed by {} brackets to write your mathematical equation or desired operation. Within the {} brackets you will need to add output() to determine the 
   2.2 When writing a for loop in R, you will need to use the () brackets to assign the amount of iterator i that you want in your dataset, columns to run your function on. You will then write the {} brackets to write the equation/model you want to run on your iterator. It is important to also write code to store the data retrieved as it will not automatically be displayed with for loops.

3. 
```{r, load in data and packages}
#Load in packages
library(tidyverse)
# Read in the CSV file
#Im not sure why but even though my working directory is set to Git it doesn't appear to want to read in Cities unless I do a full file path read.
cities_data <- read.csv("Cities.csv")
head(cities_data)
```


```{r, function}
# Define the function
haversine_dist <- function(lat1, lon1, lat2, lon2) {
# convert to radians
  rad.lat1 <- lat1 * pi/180
  rad.lon1 <- lon1 * pi/180
  rad.lat2 <- lat2 * pi/180
  rad.lon2 <- lon2 * pi/180
  
# Haversine formula
  delta_lat <- rad.lat2 - rad.lat1
  delta_lon <- rad.lon2 - rad.lon1
  a <- sin(delta_lat / 2)^2 + cos(rad.lat1) * cos(rad.lat2) * sin(delta_lon / 2)^2
  c <- 2 * asin(sqrt(a))
  
# Earth's radius in meters (standard WGS84), converted to km later
  earth_radius <- 6378137 
  
# Calculate the distance of earths radius in kilometers
  distance_km <- (earth_radius * c)/1000
  
  return(distance_km)
}
```


```{r, nyc/auburn}

# Subset the rows for New York and Auburn
nyc_row <- cities_data[cities_data$city == "New York", ]
auburn_row <- cities_data[cities_data$city == "Auburn", ]

# Extract coordinates
lat_nyc <- nyc_row$lat
long_nyc <- nyc_row$long
lat_aub <- auburn_row$lat
long_aub <- auburn_row$long

# Pass values into function
nyc_distance <- haversine_dist(lat1 = lat_aub, lon1 = long_aub, 
                               lat2 = lat_nyc, lon2 = long_nyc)

# Print result
print(nyc_distance)
```


```{r, all cities}
# Get Auburn coordinates using column names
auburn_row <- cities_data[cities_data$city == "Auburn", ]
auburn_lat <- auburn_row$lat
auburn_long <- auburn_row$long

# Initialize vector for storage
all_distances <- numeric(nrow(cities_data))

# Run Loop
for (i in 1:nrow(cities_data)) {
  current_lat <- cities_data$lat[i]
  current_long <- cities_data$long[i]
  dist <- haversine_dist(lat1 = auburn_lat, lon1 = auburn_long,
                         lat2 = current_lat, lon2 = current_long)
  all_distances[i] <- dist
  print(dist)
}
```


```{r, all cities vector transformation}
# Modify loop to append rows to a dataframe
# Create a null object to store the final results
auburn_distances_df <- NULL
# Get Auburn coordinates once
auburn_row <- cities_data[cities_data$city == "Auburn", ]
auburn_lat <- auburn_row$lat
auburn_long <- auburn_row$long

# Run the For Loop
for (i in seq_along(cities_data$city)) {
  current_city_name <- cities_data$city[i]
  current_lat <- cities_data$lat[i]
  current_long <- cities_data$long[i]
  dist <- haversine_dist(lat1 = auburn_lat, lon1 = auburn_long,
                         lat2 = current_lat, lon2 = current_long)
  city_row_i <- data.frame(
    City1 = current_city_name,
    City2 = "Auburn",
    Distance_km = dist,
    stringsAsFactors = FALSE
  )
#Append new data frame
  auburn_distances_df <- rbind.data.frame(auburn_distances_df, city_row_i)
}

# Show fist 6 rows
head(auburn_distances_df, 6)
```