---
title: "Markdown Challenge Carson"
author: "Carson Archie"
date: "2026-02-19"
output: word_document:
    md_document:
    variant: gfm
  html_document:
     toc: true
    toc_float: true
  pdf_document:
  word_document:

---

# Contents

### Coding Challenge 4
- Question 1
- Question 2
- Question 3
- Question 4
- Question 5
- Question 6

# Coding Challenge 4

## Question 1

- a. A YAML header is a block of coded data at the top of an R Markdown document with the documents title, author, date, and output type. 
- b. A pull request is when you pull a copy or "branched" version of your code in GITHub to edit without breaking your existing code and running it on your own computer. 
- c. Literate programing is when a data scientists creates a document with a tool like R Markdown to merge dictated notes, code chunks and code output into one file for viewing. 

## Question 2

Load in the library you will need.
```{r}
library(tidyverse)
library(ggplot2)
library(ggpubr)
```
Colorblind Pallet Needed
```{r}
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

### Question 2a - clickable link to manuscript where data are published 

- Noel, Z.A., Roze, L.V., Breunig, M., Trail, F. 2022. Endophytic fungi promising biocontrol agent to protect wheat from Fusarium graminearum head blight. Plant Disease. 
https://doi.org/10.1094/PDIS-06-21-1253-RE

- The following github repository is associated with this paper: 
https://github.com/noelzach/EndophyteBiocontrol

### Question 2b - Read in the data using a relative file path
```{r}
DON_data <- read.csv("/Users/carsonarchie/Desktop/MycotoxinData (1).csv", na.strings = "na") 
```

### Question 2c - 

```{r}
DON_plot <- ggplot(DON_data, aes(x = Treatment, y = DON, fill = Cultivar)) +
  geom_boxplot(outliers = T, outlier.shape = 15) + 
  geom_point(position = position_jitterdodge(), shape = 21, alpha = 0.6) + 
  ylab("DON(ppm)") + 
  xlab("") + 
  geom_jitter(pch = 21, position = position_jitterdodge(), color = "black", alpha = 0.6) +
  scale_fill_manual(values = c(cbbPalette[[3]], cbbPalette[[4]])) +
  theme_classic() +
  facet_wrap(~Cultivar)
DON_plot



```
##DON figure
```{r}
 stats_donplot <- DON_plot + geom_pwc(aes(group = Treatment), method = "t-test", label = "{p.adj.format}{p.adj.signif}", p.adjust.methods)
```

```{r}
DON_15 <- ggplot(DON_data, aes(x = Treatment, y = X15ADON, fill = Cultivar)) +
  geom_boxplot(outliers = F) + 
  geom_point(position = position_jitterdodge(), shape = 21, alpha = 0.6) + 
  ylab("DON(ppm)") + 
  xlab("") + 
  geom_jitter(pch = 21, position = position_jitterdodge(), color = "black", alpha = 0.6) +
  scale_fill_manual(values = c(cbbPalette[[3]], cbbPalette[[4]])) +
  theme_classic() +
  facet_wrap(~Cultivar)
DON_15
```

```{r}
 stats_DON_15 <- DON_15 + geom_pwc(aes(group = Treatment), method = "t-test", label = "{p.adj.format}{p.adj.signif}", p.adjust.methods)
```

```{r}
seedmass <- ggplot(DON_data, aes(x = Treatment, y = MassperSeed_mg, fill = Cultivar)) +
  geom_boxplot(outliers = F) + 
  geom_point(position = position_jitterdodge(), shape = 21, alpha = 0.6) + 
  ylab("DON(ppm)") + 
  xlab("") + 
  geom_jitter(pch = 21, position = position_jitterdodge(), color = "black", alpha = 0.6) +
  scale_fill_manual(values = c(cbbPalette[[3]], cbbPalette[[4]])) +
  theme_classic() +
  facet_wrap(~Cultivar)
seedmass
```

```{r}
 stats_seedmass <- seedmass + geom_pwc(aes(group = Treatment), method = "t-test", label = "{p.adj.format}{p.adj.signif}", p.adjust.methods)
```

```{r}
ggarrange(stats_donplot, stats_DON_15, stats_seedmass, labels = "auto", ncol = 3, nrow = 1, common.legend = TRUE) 
```
## Question 3
The Knit process is working but it is still not allowing me to puch this document to my github repostiory. 

## Question 4
I am going to The main office for IT because my current repository link isn't pushing and the college of Agricultures IT department didn't know what to do. 


## Question 5
Hopefully I will get the link set tomorrow I have an appointment with the IT department in the morning so I will send an email when my repository is updated. I feel confident with the knit functions and README file formatting but my computer is being complicated. 


## Question 6
```{r}

```